Calculating the Operator Norm in Julia

### 2025-09-10 paper/paper.qmd
rename to BlockTensorFactorization.jl

---
title: "BlockTensorDecompositions.jl: A Unified Constrained Tensor Decomposition Julia Package"
subtitle: ""
authors:
  - name: "Nicholas J. E. Richardson"
    affiliations:
      - ref: math
  - name: "Noah Marusenko"
    affiliations:
      - ref: cs
  - name: "Michael P. Friedlander"
    affiliations:
      - ref: mathcs
affiliations:
  - id: math
    name: "Department of Mathematics"
  - id: cs
    name: "Department of Computer Science"
  - id: mathcs
    name: "Departments of Mathematics and Computer Science\n"
date: ""
date-format: "D MMMM YYYY"
slide-number: true
categories: [tensors, signal processing, julia]
execute:
  freeze: true
bibliography: references.bib
csl: citationstyles/ieee-compressed-in-text-citations.csl # Citation Style
callout-appearance: minimal
callout-icon: false
format:
  typst:
    toc: true
    section-numbering: 1.1.a.i
    keep-typ: true
    mainfont: "Libertinus Serif"
editor:
  render-on-save: true
#engine: julia
#appendix-style: plain
# nocite: |
#   @firstCitation
#embed-resources: true ## uncomment when sharing HTML file
---

@@

For the scientific user, it would be most useful for there to be a single piece of software that can take as input 1) any reasonable type of factorization model and 2) constraints on the individual factors, and produce a factorization. Details like what rank to select, how the constraints should be enforced, and convergence criteria should be handled automatically, but customizable to the knowledgable user. These are the core specification for BlockTensorDecompositions.jl.

@@

The main contribution is a description of a fast and flexible tensor decomposition package, along with a public implementation written in Julia: BlockTensorDecompositions.jl. This package provides a framework for creating and performing custom tensor decompositions. To the author's knowledge, it is the first package to provide automatic factorization to a large class of constrained tensor decompositions problems, as well as a framework for implementing new constraints and iterative algorithms. This paper also describes three new techniques not found in the literature that empirically convergence faster than traditional block-coordinate descent.

@@

The $n$-fibres, $n$th mode fibres, or mode $n$ fibres of an $N$th order tensor $A$ are denoted $A[i_1,~\dots,~i_{n-1},~:,~i_{n+1},~\dots,~i_{N}]$. For example, the 1-fibres of a matrix $M$ are the column vectors\
$M[:,~j]$, and the 2-fibres are the row vectors $M[i,~:]$. For order-3 tensors, the $1$st, $2$nd, and $3$rd mode fibres $A[:, j, k]$, $A[i,:,:]$, and $A[i,j,:]$ are called the vertical/column, horizontal/row, and depth/tube fibres respectively and are displayed in @fig-tensor-fibres. Natively in Julia, the 1-, 2-, and 3-fibres of a third order array `A` would be `eachslice(A, dims=(2,3))`, `eachslice(A, dims=(1,3))`, and `eachslice(A, dims=(1,2))`. BlockTensorDecomposition.jl defines the function `eachfibre(A; n)` to do exactly this. For example, the 1-fibres of an array `A` would be `eachfibre(A, n=1)`.

@@

BlockTensorDecomposition.jl defines `identity_tensor(I, ndims)` to construct $\mathrm{id}_I$.

@@

TODO Define in BlockTensorDecomposition.jl

@@

This product can also be thought of as taking the dot product $(A \cdot_n B)[i, j] = A_i \cdot B_j$ between all pairs of $n$th order slices of $A$ and $B$, which exactly how BlockTensorDecomposition.jl defines the operation.

@@

BlockTensorDecomposition.jl defines this operation with `slicewise_dot(A, B, n)`. In the special case where $A=B$, a more efficient method that only computes entries where $i \leq j$ is defined since $A \cdot_n A$ is a symmetric matrix.

@@

BlockTensorDecomposition.jl defines this operation with `nmode_product(A, B, n)`.

@@

Computationally, we can think of a generic decomposition as storing factors $(A,B,C,...)$ and operations $(\times_a,\times_b,...)$ for combining them. This is what we do in BlockTensorDecomposition.jl.

@@

The *Tucker Product* defined by @eq-tucker-product is implemented in BlockTensorDecomposition.jl with `tuckerproduct(B, (A1, ..., AN))` and computes
$$
B \bigtimes_{n} A_n = \llbracket B; A_1,\dots,A_N \rrbracket.
$$

@@

There are implemented in BlockTensorDecomposition.jl and can be called, for a third order tensor, with `Tucker((B, A₁, A₂, A₃))`, `Tucker1((B, A₁))`, and `CPDecomposition((A₁, A₂, A₃))`. These Julia `structs` store the tensor in its factored form. We could define the contractions for these types and use the common interface provided by `array`, but it turns out we can reconstruct the whole tensor more efficiently. If the recombined tensor or particular entries are requested, Julia dispatches on the type of decomposition and calls a particular method of `array` or `getindex`. The implementations for efficient array construction and index access are provided below.

@@

To ensure the code stays flexible, the main algorithm of BlockTensorDecomposition.jl, `factorize`, is defined at a very high level.

@@

The associated implementation with BlockTensorDecomposition.jl is shown below. We define a `make_gradient` which takes the decomposition, factor index `n`, and data tensor `Y`, and creates a function that computes the gradient for the same type of decomposition. This lets us manipulate the function that computes the gradient, rather than just the computed gradient.

@@

Two more techniques are implemented in BlockTensorDecomposition.jl to improve convergence. To the authors knowledge, these are new to tensor factorization, but may or may not be applicable depending on the exact factorization problem or data being studied. For these reasons, these other techniques are discussed separately in @sec-ppr and @sec-multi-scale.

@@

```julia
using BenchmarkTools
using BlockTensorDecomposition

fact = BlockTensorDecomposition.factorize

@@

There are a number of software engineering techniques used to ensure BlockTensorDecomposition.jl is flexible and applicable to a wide range of problems. These enable key algorithmic choices to be hot-swapped and easily compared with each other.

@@

In finite precision, we can halt the algorithm if we can guarantee the solution is accurate to machine precision. This can often be too strict if convergence is not at a fast enough rate. Furthermore, depending on *why* we are decomposing a tensor, we may want different stats to be within a given a tolerance. BlockTensorDecomposition.jl attempts to solve this issue by defining some standard criteria that can be used to halt the algorithm. These are subtypes of the abstract type `AbstractStat` and are listed below.

@@

One of the main motivations for developing BlockTensorDecomposition.jl is to solve constrained tensor problems. Other code did not have the expressivity to handle constraints beyond the most common constraints on the factors: nonnegativity and Euclidean normalized columns. To enable flexibility, BlockTensorDecomposition.jl defines

@@

These operations agree for the Frobenius norm (entry-wise $2$-norm), but are different operations in general. In BlockTensorDecomposition.jl, we define these classes as the following.

@@

The $\max(0, x)$ function should be understood as operating entrywise on $x$. In BlockTensorDecomposition, we use the helper `ReLU(x) = max(0, x)` for this function to assist with broadcasting.

@@

We will refer to @eq-nnpr as nonnegative projection and rescaling (NNPR). NNPR has the following implementation in BlockTensorDecomposition.jl.

@@

This algorithm can be called in BlockTensorDecomposition.jl with the following code.

@@

The two algorithms can be called in BlockTensorDecomposition.jl with the following code.

@@

The relevant call in BlockTensorDecomposition would be the following.

@@

and is called in BlockTensorDecomposition.jl with the following code.

@@

The BlockTensorDecomposition.jl Julia provides a new all-in-one package for performing constrained tensor factorizations, and a playground for designing new decompositions and custom constraints. Careful design elements were engineered to balance flexibility and efficiency. By creating this package, new advancements like enforcing constraints through scaling rather than projection, and performing optimization over multiple scales were mathematically examined and numerically tested. These novel ideas are worth investigating further to see their applicability to continuous optimization beyond tensor factorization.

>>>>

---
title: "BlockTensorFactorizations.jl: A Unified Constrained Tensor Decomposition Julia Package"
subtitle: ""
authors:
  - name: "Nicholas J. E. Richardson"
    affiliations:
      - ref: math
  - name: "Noah Marusenko"
    affiliations:
      - ref: cs
  - name: "Michael P. Friedlander"
    affiliations:
      - ref: mathcs
affiliations:
  - id: math
    name: "Department of Mathematics"
  - id: cs
    name: "Department of Computer Science"
  - id: mathcs
    name: "Departments of Mathematics and Computer Science\n"
date: ""
date-format: "D MMMM YYYY"
slide-number: true
categories: [tensors, signal processing, julia]
execute:
  freeze: true
bibliography: references.bib
csl: citationstyles/ieee-compressed-in-text-citations.csl # Citation Style
callout-appearance: minimal
callout-icon: false
format:
  typst:
    toc: true
    section-numbering: 1.1.a.i
    keep-typ: true
    mainfont: "Libertinus Serif"
editor:
  render-on-save: true
#engine: julia
#appendix-style: plain
# nocite: |
#   @firstCitation
#embed-resources: true ## uncomment when sharing HTML file
---

@@

For the scientific user, it would be most useful for there to be a single piece of software that can take as input 1) any reasonable type of factorization model and 2) constraints on the individual factors, and produce a factorization. Details like what rank to select, how the constraints should be enforced, and convergence criteria should be handled automatically, but customizable to the knowledgable user. These are the core specification for BlockTensorFactorizations.jl.

@@

The main contribution is a description of a fast and flexible tensor decomposition package, along with a public implementation written in Julia: BlockTensorFactorizations.jl. This package provides a framework for creating and performing custom tensor decompositions. To the author's knowledge, it is the first package to provide automatic factorization to a large class of constrained tensor decompositions problems, as well as a framework for implementing new constraints and iterative algorithms. This paper also describes three new techniques not found in the literature that empirically convergence faster than traditional block-coordinate descent.

@@

The $n$-fibres, $n$th mode fibres, or mode $n$ fibres of an $N$th order tensor $A$ are denoted $A[i_1,~\dots,~i_{n-1},~:,~i_{n+1},~\dots,~i_{N}]$. For example, the 1-fibres of a matrix $M$ are the column vectors\
$M[:,~j]$, and the 2-fibres are the row vectors $M[i,~:]$. For order-3 tensors, the $1$st, $2$nd, and $3$rd mode fibres $A[:, j, k]$, $A[i,:,:]$, and $A[i,j,:]$ are called the vertical/column, horizontal/row, and depth/tube fibres respectively and are displayed in @fig-tensor-fibres. Natively in Julia, the 1-, 2-, and 3-fibres of a third order array `A` would be `eachslice(A, dims=(2,3))`, `eachslice(A, dims=(1,3))`, and `eachslice(A, dims=(1,2))`. BlockTensorFactorization.jl defines the function `eachfibre(A; n)` to do exactly this. For example, the 1-fibres of an array `A` would be `eachfibre(A, n=1)`.

@@

BlockTensorFactorization.jl defines `identity_tensor(I, ndims)` to construct $\mathrm{id}_I$.

@@

TODO Define in BlockTensorFactorization.jl

@@

This product can also be thought of as taking the dot product $(A \cdot_n B)[i, j] = A_i \cdot B_j$ between all pairs of $n$th order slices of $A$ and $B$, which exactly how BlockTensorFactorization.jl defines the operation.

@@

BlockTensorFactorization.jl defines this operation with `slicewise_dot(A, B, n)`. In the special case where $A=B$, a more efficient method that only computes entries where $i \leq j$ is defined since $A \cdot_n A$ is a symmetric matrix.

@@

BlockTensorFactorization.jl defines this operation with `nmode_product(A, B, n)`.

@@

Computationally, we can think of a generic decomposition as storing factors $(A,B,C,...)$ and operations $(\times_a,\times_b,...)$ for combining them. This is what we do in BlockTensorFactorization.jl.

@@

The *Tucker Product* defined by @eq-tucker-product is implemented in BlockTensorFactorization.jl with `tuckerproduct(B, (A1, ..., AN))` and computes
$$
B \bigtimes_{n} A_n = \llbracket B; A_1,\dots,A_N \rrbracket.
$$

@@

There are implemented in BlockTensorFactorization.jl and can be called, for a third order tensor, with `Tucker((B, A₁, A₂, A₃))`, `Tucker1((B, A₁))`, and `CPDecomposition((A₁, A₂, A₃))`. These Julia `structs` store the tensor in its factored form. We could define the contractions for these types and use the common interface provided by `array`, but it turns out we can reconstruct the whole tensor more efficiently. If the recombined tensor or particular entries are requested, Julia dispatches on the type of decomposition and calls a particular method of `array` or `getindex`. The implementations for efficient array construction and index access are provided below.

@@

To ensure the code stays flexible, the main algorithm of BlockTensorFactorization.jl, `factorize`, is defined at a very high level.

@@

The associated implementation with BlockTensorFactorization.jl is shown below. We define a `make_gradient` which takes the decomposition, factor index `n`, and data tensor `Y`, and creates a function that computes the gradient for the same type of decomposition. This lets us manipulate the function that computes the gradient, rather than just the computed gradient.

@@

Two more techniques are implemented in BlockTensorFactorization.jl to improve convergence. To the authors knowledge, these are new to tensor factorization, but may or may not be applicable depending on the exact factorization problem or data being studied. For these reasons, these other techniques are discussed separately in @sec-ppr and @sec-multi-scale.

@@

```julia
using BenchmarkTools
using BlockTensorFactorization

fact = BlockTensorFactorization.factorize

@@

There are a number of software engineering techniques used to ensure BlockTensorFactorization.jl is flexible and applicable to a wide range of problems. These enable key algorithmic choices to be hot-swapped and easily compared with each other.

@@

In finite precision, we can halt the algorithm if we can guarantee the solution is accurate to machine precision. This can often be too strict if convergence is not at a fast enough rate. Furthermore, depending on *why* we are decomposing a tensor, we may want different stats to be within a given a tolerance. BlockTensorFactorization.jl attempts to solve this issue by defining some standard criteria that can be used to halt the algorithm. These are subtypes of the abstract type `AbstractStat` and are listed below.

@@

One of the main motivations for developing BlockTensorFactorization.jl is to solve constrained tensor problems. Other code did not have the expressivity to handle constraints beyond the most common constraints on the factors: nonnegativity and Euclidean normalized columns. To enable flexibility, BlockTensorFactorization.jl defines

@@

These operations agree for the Frobenius norm (entry-wise $2$-norm), but are different operations in general. In BlockTensorFactorization.jl, we define these classes as the following.

@@

The $\max(0, x)$ function should be understood as operating entrywise on $x$. In BlockTensorFactorization, we use the helper `ReLU(x) = max(0, x)` for this function to assist with broadcasting.

@@

We will refer to @eq-nnpr as nonnegative projection and rescaling (NNPR). NNPR has the following implementation in BlockTensorFactorization.jl.

@@

This algorithm can be called in BlockTensorFactorization.jl with the following code.

@@

The two algorithms can be called in BlockTensorFactorization.jl with the following code.

@@

The relevant call in BlockTensorFactorization would be the following.

@@

and is called in BlockTensorFactorization.jl with the following code.

@@

The BlockTensorFactorization.jl Julia provides a new all-in-one package for performing constrained tensor factorizations, and a playground for designing new decompositions and custom constraints. Careful design elements were engineered to balance flexibility and efficiency. By creating this package, new advancements like enforcing constraints through scaling rather than projection, and performing optimization over multiple scales were mathematically examined and numerically tested. These novel ideas are worth investigating further to see their applicability to continuous optimization beyond tensor factorization.

### 2025-09-10 test/opnorm/faster_opnorm.qmd
rename to BlockTensorFactorization.jl

## Possible routines
```{julia}
using Pkg; Pkg.activate(".")
Pkg.add(["LinearAlgebra","Random","Arpack","BenchmarkTools"])
#Pkg.add(path="C:\\Users\\Nicholas\\OneDrive - UBC\\Research\\general-block-decomposition\\BlockTensorDecomposition.jl")
```
```{julia}
using Random: rand, randn
using LinearAlgebra: svdvals, opnorm, Symmetric
using Arpack: svds
using BlockTensorDecomposition: slicewise_dot
using BenchmarkTools
```

>>>>

## Possible routines
```{julia}
using Pkg; Pkg.activate(".")
Pkg.add(["LinearAlgebra","Random","Arpack","BenchmarkTools"])
#Pkg.add(path="C:\\Users\\Nicholas\\OneDrive - UBC\\Research\\general-block-decomposition\\BlockTensorFactorization.jl")
```
```{julia}
using Random: rand, randn
using LinearAlgebra: svdvals, opnorm, Symmetric
using Arpack: svds
using BlockTensorFactorization: slicewise_dot
using BenchmarkTools
```
